### 2025-08-12 chapters/background.tex
updated TOC
added baseline results

\chapter{Background}
\label{chap:background}

@@

\section{Keystone Architecture Overview}

@@

\section{Kyber: A Post-Quantum Cryptography Algorithm}
\label{sec:kyber}

@@

@@

\section{Comparison with Other TEEs}
\label{sec:comparison_tees}

@@

\section{Performance Considerations in TEEs}

>>>>

\chapter{Background and Related Technologies}
\label{chap:background}

@@

\section{Keystone Enclave Architecture}

@@

\section{Physical Memory Protection (PMP) in RISC-V}
\label{sec:pmp}

Physical Memory Protection (PMP) is a hardware-level security mechanism defined by the RISC-V privileged architecture to provide fine-grained control over access permissions to physical memory regions. PMP plays a central role in enabling secure system software, particularly in contexts where memory isolation between privileged and unprivileged software is essential. It is a critical enabler for implementing secure environments such as hypervisors, operating system kernels, and Trusted Execution Environments (TEEs) without relying on traditional virtualization support~\cite{riscvprivspec}.

In RISC-V, the PMP mechanism is configured exclusively by machine-mode (M-mode) software—the most privileged execution level in the architecture—and applies access control policies to all lower-privileged software running in supervisor (S-mode) or user mode (U-mode). Each PMP entry defines a memory region along with a set of permissions (read, write, execute) and an addressing mode. The hardware enforces these permissions such that any access attempt by S-mode or U-mode software that violates a PMP rule results in a trap to M-mode. This mechanism ensures that M-mode retains full control over physical memory while being able to delegate controlled access to untrusted software layers~\cite{riscvprivspec,lee2019keystone}.

A typical RISC-V core supports a fixed number of PMP entries (commonly 8 or 16), each represented by two Control and Status Registers (CSRs): \texttt{pmpcfg} and \texttt{pmpaddr}. The \texttt{pmpcfg} register specifies the access permissions and address matching mode for each region, while \texttt{pmpaddr} defines the address bounds. RISC-V supports several region encoding modes:

\begin{itemize}
    \item \textbf{TOR (Top-of-Range):} Defines a region from the base address of the previous PMP entry up to the current address.
    \item \textbf{NAPOT (Naturally Aligned Power-Of-Two):} Encodes both the base and size in a single register, enabling efficient representation of power-of-two-sized regions.
    \item \textbf{NA4 (Naturally Aligned 4-byte):} Encodes a minimal 4-byte memory region.
\end{itemize}

The design of PMP aligns with the minimalist and modular philosophy of RISC-V. Unlike traditional memory protection mechanisms such as x86's segmentation or ARM's TrustZone Address Space Controller (TZASC), PMP is simple, deterministic, and free from microarchitectural complexity. Importantly, PMP operates at the physical memory level—independent of virtual memory or page tables—which makes it highly suitable for systems with limited resources or without Memory Management Units (MMUs)~\cite{riscvprivspec,Survey2023}.

In the context of TEEs such as the Keystone framework, PMP serves as the fundamental enforcement mechanism for enclave isolation. Keystone leverages PMP to assign a protected memory region, known as the Enclave Page Memory (EPM), to each enclave. This region is configured to be inaccessible to both the untrusted operating system (running in S-mode) and any other software outside the enclave context~\cite{dayeol2019keystone}. During enclave creation, the Keystone Security Monitor (SM)—executing in M-mode—programmatically sets PMP entries to lock the memory region associated with the enclave. The SM verifies the integrity of the enclave binary, computes its cryptographic hash, and sets up the memory layout in such a way that only the enclave code has access to its memory contents~\cite{lee2019keystone,keystone2025how}.

Furthermore, PMP entries are reconfigured dynamically during enclave context switches. For example, when the host operating system regains control, the PMP settings are altered to revoke access to the enclave’s memory. Conversely, when the enclave resumes execution, the SM reinstates its PMP region. This dynamic switching is essential to enforce temporal isolation and prevent time-of-check-to-time-of-use (TOCTTOU) vulnerabilities that may arise during state transitions~\cite{cerberus2022}.

Despite its utility, PMP has several practical limitations. The fixed number of PMP entries restricts the number of simultaneously protected regions, posing a challenge for complex systems with many isolated domains or subregions. Moreover, overlapping PMP regions are resolved based on entry priority (lower-numbered entries take precedence), requiring careful configuration to avoid inadvertent leaks or privilege escalations~\cite{riscvprivspec}.

Security researchers have explored various enhancements and formal models to increase the reliability and verifiability of PMP-based isolation. For example, Cerberus~\cite{cerberus2022} builds a formally verified memory-sharing abstraction over PMP to enable safe communication between enclaves and untrusted components. Similarly, HexBox~\cite{hexbox2018} leverages PMP for real-time embedded security by associating memory regions with specific control-flow paths in embedded firmware, demonstrating PMP's applicability beyond general-purpose computing.

The role of PMP in Keystone exemplifies how low-complexity hardware primitives can be orchestrated by secure software layers to implement strong isolation guarantees. This synergy is particularly effective in the RISC-V ecosystem, where the open architecture allows deep customization, formal verification, and academic experimentation~\cite{Survey2023}. The combination of PMP, a secure boot mechanism, and a minimal TCB enables Keystone to enforce hardware-rooted enclave isolation without relying on opaque, vendor-specific extensions, setting it apart from alternatives like Intel SGX or ARM TrustZone~\cite{dayeol2019keystone,turn0search0}.

In conclusion, Physical Memory Protection in RISC-V is a powerful, lightweight, and deterministic mechanism that underpins secure execution in many RISC-V-based platforms. Its integration into frameworks like Keystone enables transparent and verifiable enclave isolation while remaining suitable for constrained systems. As the RISC-V ecosystem continues to evolve, PMP will remain a fundamental building block for secure, open, and formally analyzable trusted computing platforms.

\section{Post-Quantum Cryptography: Kyber Algorithm}
\label{sec:kyber}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=0.9\linewidth]{figures/kyber_768.png}
\caption{Kyber768 Key Exchange Algorithm}
\label{fig:kyber_768}
\end{figure}

@@

\section{Comparative Overview of TEE}
\label{sec:comparison_tees}

@@

\section{Performance Challenges in Secure Execution}

### 2025-08-12 chapters/benchmark.tex
added baseline results

\section{Baseline: Native (Non-Enclave) Performance}

To establish a baseline for evaluating enclave overheads, we first executed all benchmarks natively on the simulated RISC-V platform without any enclave isolation. This phase allows us to measure the raw performance characteristics of the system and isolate the costs introduced solely by secure execution.

The Dhrystone and CoreMark benchmarks exhibited expected scaling behavior with respect to CPU cores and memory size. As CPU core count increased, throughput for these parallelizable benchmarks improved approximately linearly, reflecting efficient multi-threaded execution on the underlying hardware. Memory size impacted CoreMark more noticeably, particularly during memory-intensive phases, where increased memory reduced paging and improved cache hit rates. In contrast, Dhrystone’s smaller working set resulted in negligible sensitivity to memory size variations.

The Kyber cryptographic benchmark behaved differently. Due to its inherently sequential algorithmic structure, Kyber’s execution time remained relatively constant across changes in CPU core count. Likewise, memory size had minimal impact on Kyber performance, as its working set comfortably fit within allocated memory without inducing paging.

These native baseline results confirm the stability and predictability of the simulated system environment, providing a solid foundation for subsequent enclave overhead analysis.

>>>>

\section{Baseline: Native (Non-Enclave) Performance}
\label{sec:baseline-native}

To establish a reliable baseline for evaluating the performance overhead introduced by secure enclaves, we first measured native (non-enclave) execution of key benchmarks on the simulated RISC-V platform. All benchmarks were executed under controlled conditions consistent with the previously described parameter variation framework, focusing primarily on the Low-End Baseline setup with 64~MB RAM, a single CPU core, and sequential execution.

The selected benchmarks include synthetic workloads CoreMark and Dhrystone, both widely used in embedded systems performance evaluation, as well as cryptographic primitives from the Kyber post-quantum key encapsulation mechanism (KEM). Each benchmark was run multiple times---typically ten iterations---with the initial run discarded to mitigate artifacts from JIT warm-up and caching effects inherent to the QEMU TCG runtime. Reported metrics represent averages over the remaining nine runs, with standard deviations provided where applicable.

Key performance metrics collected include wall-clock execution time (measured using high-resolution timers), throughput indicators such as iterations or Dhrystones per second, CPU utilization (user and system time), memory footprint (peak resident set size), page fault counts, context switch statistics, and cache metrics obtained via the QEMU TCG cache plugin.

Figure~\ref{fig:coremark_native} presents CoreMark results under the 10-run sequential configuration. The benchmark demonstrated high stability, with an average elapsed time of 107.5 seconds and approximately 939 iterations per second. CPU utilization was predominantly user-space (81.4 seconds) with minimal system overhead (1 second). Memory consumption remained low at 10~KB RSS, and page fault and context switch counts indicated a minimally interrupted environment.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/coremark_64M_1Core_native_metrics.pdf}
    \caption{CoreMark Native Execution Summary (10 Runs, Low-End Baseline)}
    \label{fig:coremark_native}
\end{figure}

Figure~\ref{fig:dhrystone_native} summarizes Dhrystone benchmark native execution results. The mean elapsed time was approximately 102.8 seconds, with throughput near 96,200 Dhrystones per second. User CPU time was steady at 10.3 seconds, and system CPU time was negligible. Minor page faults were more frequent than in CoreMark, reflecting normal demand paging, and involuntary context switches occurred regularly due to scheduler activity but did not noticeably impact throughput.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/dhrystone_64M_1Core_native_metrics.pdf}
    \caption{Dhrystone Native Execution Summary (10 Runs, Low-End Baseline)}
    \label{fig:dhrystone_native}
\end{figure}

Figure~\ref{fig:kyber_native_cycles} reports median CPU cycle counts for Kyber cryptographic operations measured in native mode on the Low-End Baseline configuration. Keypair generation, encapsulation, and decapsulation were profiled across Kyber-512, Kyber-768, and Kyber-1024 parameter sets. All showed consistent cycle counts, indicating stable execution and resource usage.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/kyber_64M_1Core.pdf}
    \caption{Kyber Native Execution: Average CPU Cycles per Operation}
    \label{fig:kyber_native_cycles}
\end{figure}

Cache statistics collected via the QEMU plugin across all benchmarks indicated predictable access patterns consistent with compute-bound workloads. Instruction and data cache miss rates remained low, reinforcing the stability of native execution on this platform.

In summary, these native baseline results confirm a stable and well-characterized environment with minimal interference or resource contention. This baseline provides a critical reference for comparing enclave-mode execution and precisely quantifying the overhead of secure execution under representative embedded system constraints.
